---
title: "assigment Session 6"
author:
  name: Jan Pamula
format: 
  html:
      embed-resources: true
      theme: lumen
      toc: true
      toc-color: red
      toc-location: left
      toc-depth: 6
      toc-expand: 6
      code-tools:
        source: true
        toggle: true
      code-fold: true
      code-summary: "Show the code"
      code-overflow: wrap
      page-layout: full
editor: visual
editor_options: 
  markdown: 
    wrap: 72
---

# 1- Load the "iris" dataset.

Loading libraries

```{r}

library(tidyr)
library(dplyr)

```

Loading "iris" dataset

```{r}

data("iris")

```

# 2- Transform it to longer

```{r}

iris_long<- iris %>% 
  tidyr::pivot_longer(cols=matches("Sepal.Length|Sepal.Width|Petal.Length|Petal.Width"), names_to = "Trait", values_to = "Size"
                      
```

transforming "iris" database to "longer" by function
`tidyr::pivot_longer`:

-   `iris_long<-` - name of new dataset

-   `%>%` - symbol for concatenate operations

-   `cols=matches("Sepal.Length|Sepal.Width|Petal.Length|Petal.Width")` -
    indicate columns we want to pivot

-   `names_to = "Trait"` - how column witch names of pivoting columns
    (Sepal.Length, Sepal.Width, Petal.Length, Petal.Width) will be
    called ("Trait")

-   `values_to = "Size"` - how will be column called ("Size") with
    values of pivoting columns

# 3- Answer the following questions.

## A) What is the mean length and width of the sepal and petal of each species?

```{r}

iris_mean <- iris_long %>% 
  group_by(Species, Trait) %>% 
  summarise(MeanSize = mean(Size))

```

-   `iris_mean <-` - name of new dataset

-   `%>%` - symbol for concatenate operations

-   `group_by(Species, Trait)` - grouping by columns

-   `summarise(MeanSize = mean(Size))` - creating new dataset with
    grouped by columns and new column (named "MeanSize") which is equal
    mean of specific group

## B) What is the largest and widest, petal and sepal, of virginica and versicolor species?

```{r}

largest_widest_virginica_versicolor <- iris_long %>%
  dplyr::filter(Species %in% c("virginica", "versicolor")) %>%
  group_by(Species, Trait) %>% 
  summarise(MaxSize = max(Size))

```

-   `largest_widest_virginica_versicolor <-` - name of new dataset

-   `%>%` - symbol for concatenate operations

-   `dplyr::filter(Species %in% c("virginica", "versicolor"))` - filter
    for records that iclude "virginica" or "versicolor" in column
    "Species"

-   `group_by(Species, Trait)` - grouping by columns

-   `summarise(MaxSize = max(Size))` - creating new dataset with grouped
    by columns and new column (named "MaxSize") which is equal max value
    of specific group

## C) For each species, for sepal and petal, width and length, how many observations are above the mean?

```{r}

over_mean <- iris_long  %>%
  filter(Size > mean(Size)) %>% 
  group_by(Species, Trait) %>% 
  summarise(n = n())

```

-   `over_mean <-` - name of new dataset

-   `%>%` - symbol for concatenate operations

-   `filter(Size > mean(Size))` - filter rows where Size \> mean(Size)

-   `group_by(Species, Trait)` - grouping by columns

-   `summarise(n = n())` - creating new dataset with grouped by columns
    and new column (named "n") which is equal number of appearances of
    specific group
